A generalized definition of multidimensional item response theory parameters

### 2022-11-18 output/paper_draft/paper_M2PLM_parameters.Rmd
Completes section "Linear algebra of non-orthogonal coordinates"

\begin{equation}
  `r DIR_COSINE_PROOF_EQ`,                        (\#eq:dir-cosine-proof-prop-1)
\end{equation}

@@

From Equation \@ref(eq:direction-cosine-vector),
any trait vector $`r TRAIT_VECTOR`$ is expressed in polar coordinates as

Thus, the M2PL model is expressed in polar coordinates as

@@

Following the procedure outlined by @reckase_difficulty_1985
[@reckase_discriminating_1991],
we first compute the point of maximum slope by finding the root(s) of
the second derivative with respect to $`r TRAIT_NORM`$.

@@

The root we are interested in is found for $`r IRF_MAX_SLOPE_EQ`$
(the other ones occur when $`r IRF_ABBR`$ equals 0 and 1,
which result in improper values of $`r TRAIT_NORM`$).
The slope in direction $`r DIR_ANGLE_VEC`$ is given by

\begin{equation}
  `r IRF_1ST_DIFF_EQ`.                                      (\#eq:M2PL-1st-diff)
\end{equation}

When $`r IRF_MAX_SLOPE_EQ`$,

@@

Notation aside, these equations differ from
Reckase's [-@reckase_difficulty_1985] Equations 3 through 6
only in the term $`r INNER_PROD_INV_DIAG_SQ_PROD`$.
When $`r LS_BASIS`$ is the standard basis,
this term simplifies to an identity matrix
(which results in Equation \@ref(eq:M2PL-formula-polar) through
\@ref(eq:slope-max) simplyfing to rectangular Cartesian coordinates,
i.e. the equations outlined by Reckase).
To compute the direction of the maximum slope,
Reckase leverages on the property that the squared direction cosines sum up to 1
(see Equation 7 therein).
However, this assertion implicitly assumes orthogonal coordinates,
so it does not apply in general to $`r DIR_COS_VEC`$.<!-- # TODO: To appendix the following, or maybe give for granted? -->
Instead, we may consider the direction cosine vector in $`r LATENT_SPACE_STD`$:

\begin{equation}
  `r DIR_COS_VEC_STD_EQ`.                     (\#eq:direction-cosine-vector-std)
\end{equation}

Operating on Equation \@ref(eq:trait-vector-polar),

\begin{equation}
  \begin{split}
    `r DIR_COS_AS_DIR_COS_STD_EQ` \\
    `r TRAIT_POLAR_COEFF_COS_STD_EQ`
  \end{split}                                 (\#eq:direction-cosine-vectors-eq)
\end{equation}

and, substituting in \@ref(eq:slope-max),

@@

We can define

\begin{equation}
  `r DISCR_VECTOR_STD_EQ`.                               (\#eq:discr-vector-std)
\end{equation}

such that Equation \@ref(eq:slope-max-std-cosines) is similar
(in vectorial form) to Reckase's [-@reckase_difficulty_1985] Equation 6,
and thus can be solved by the same procedure
applying their Equations 7 through 9.
The resulting equivalent to their Equation 9, in vectorial form
and solving for $`r DIR_ANGLE_STD_SUBSTITUTE`$ is

\begin{equation}
  `r DIR_COS_STD_ITEM_VEC_EQ`.                            (\#eq:dir-cos-vec-std)
\end{equation}

Substituting $`r DIR_COS_STD_ITEM_VEC`$ and $`r DISCR_VECTOR_STD`$
by their respective expressions results in the expression for
the direction cosine vector of item $i$ in $`r LATENT_SPACE`$,

@@

which gives the direction from the origin to the point where
the slope is maximum.
Finally, to determine the signed distance $`r DISTANCE_PARAM`$
from the origin to that point,
we solve Equation \@ref(eq:M2PL-formula-polar) for $`r IRF_MAX_SLOPE_EQ`$,
$`r DIR_COS_SUBSTITUTE`$ and
$`r DISTANCE_SUBSTITUTE`$,
and then substitute Equation \@ref(eq:dir-cos-vec) for $`r DIR_COS_ITEM_VEC`$.
The result, solving for $`r DISTANCE_PARAM`$, is

@@

The slope $`r SLOPE_MAX_PARAM`$ at the point defined by Equations
\@ref(eq:dir-cos-vec) and \@ref(eq:distance-param)
is given by Equation \@ref(eq:slope-max).
Solving for $`r DIR_COS_SUBSTITUTE`$
and substituting Equation \@ref(eq:dir-cos-vec) for $`r DIR_COS_ITEM_VEC`$
[@reckase_discriminating_1991],

>>>>

\begin{equation}
  `r DIR_COSINE_PROOF_EQ`,
\end{equation}

@@

From Proposition 1,
any trait vector $`r TRAIT_VECTOR`$ is expressed in polar coordinates as

\begin{equation}
  `r TRAIT_VECTOR_POLAR_EQ`.                      (\#eq:trait-equivalence-polar)
\end{equation}

Thus, the M2PL model is expressed in polar coordinates as
<!-- # TODO: Use the "module" symbol instead of the "norm"? -->

@@

Following the procedure outlined by @reckase_difficulty_1985
[@reckase_discriminating_1991],
we first compute the point of maximum slope by finding the root(s) of
the second derivative with respect to $`r TRAIT_NORM`$.
<!-- # TODO: Use the "module" symbol instead of the "norm"? -->

The slope in direction $`r DIR_ANGLE_VEC`$ is given by

\begin{equation}
  `r IRF_1ST_DIFF_EQ`,                                      (\#eq:M2PL-1st-diff)
\end{equation}

@@

The root we are interested in is found for $`r IRF_MAX_SLOPE_EQ`$
(the other ones occur when $`r IRF_ABBR`$ equals 0 and 1,
which result in improper values of $`r TRAIT_NORM`$).

The slope in direction $`r DIR_ANGLE_VEC`$ when $`r IRF_MAX_SLOPE_EQ`$ is given
by

@@

Notation aside, these equations differ from
Reckase's [-@reckase_difficulty_1985] Equations 3 through 6
only in the term $`r INNER_PROD_INV_DIAG_SQ_PROD`$.

To compute the direction of the maximum slope,
Reckase leverages on the property that the squared direction cosines sum up to 1
(see Equation 7 therein).
However, this assertion implicitly assumes orthogonal coordinates,
so it does not apply in general to $`r DIR_COS_VEC`$.
Instead, we may consider the direction cosine vector with the standard basis
$`r LS_STD_BASIS`$.<!-- # TODO: Mario wrote "\mathcal{B}", I think he refers to "\mathcal{E}", the standard basis -->
Now, applying Proposition 2 (with $`r ALT_MATRIX_AS_ID`$) we obtain

\begin{equation}
  \begin{split}
    `r DIR_COS_AS_DIR_COS_STD_EQ` \\
    `r TRAIT_POLAR_COEFF_COS_STD_EQ`.
  \end{split}                                 (\#eq:direction-cosine-vectors-eq)
\end{equation}

Therefore, substituting in \@ref(eq:slope-max),

@@

We can then define

\begin{equation}
  `r DISCR_VECTOR_STD_EQ`                                (\#eq:discr-vector-std)
\end{equation}

so

\begin{equation}
  `r SLOPE_MAX_STD_EQ`.                                     (\#eq:slope-max-std)
\end{equation}

We are thus considering the direction cosines with respect to the orthogonal
axes, so we can just apply @reckase_difficulty_1985's results, obtaining

\begin{equation}
  `r DIR_COS_VEC_STD_EQ`.                                 (\#eq:dir-cos-vec-std)
\end{equation}

Substituting $`r DIR_COS_VEC_STD`$
(Equation \@ref(eq:direction-cosine-vectors-eq)) and
$`r DISCR_VECTOR_STD`$ (Equation \@ref(eq:discr-vector-std))
by their respective expressions results in the
expression for the direction cosine vector of item $i$ in $`r LATENT_SPACE`$,
<!-- # TODO: with basis \mathcal{B}? -->

@@

which gives the direction from the origin to the point where
the slope is maximum.
Finally, to determine the signed distance $`r DISTANCE_PARAM`$
from the origin to that point,
we solve Equation \@ref(eq:M2PL-formula-polar) for $`r IRF_MAX_SLOPE_EQ`$ to get
$`r TRAIT_NORM_SOLVED_EQ`$. Using equation \@ref(eq:dir-cos-vec), then
<!-- # TODO: Verify that it's equation 20, not 18 -->

@@

The slope $`r SLOPE_MAX_PARAM`$ at the point defined by Equations
\@ref(eq:dir-cos-vec) and \@ref(eq:distance-param)
is given by Equation \@ref(eq:slope-max).
Substituting Equation \@ref(eq:dir-cos-vec) in Equation \@ref(eq:slope-max),
[@reckase_discriminating_1991],
